---
title: "Midterm project"
author: "Carmen Chen"
date: "10/20/2021"
output: 
  github_document: default
  html_document:
    html_preview: false
  word_document: default
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(tidyverse)
```

Introduction (provide background on your dataset and formulated question)

Vaccination rate and COVID-19 incidence cases in California.

```{r read in the data}
cases <- read.csv("Cases.csv")
vaccinations <- read.csv("Vaccinations.csv")
```

```{r clean the data}
table(vaccinations$Location)
vaccinations <- vaccinations[vaccinations$Location == "CA", ]
```


Methods (include how and where the data were acquired, how you cleaned and wrangled the data, what tools you used for data exploration)

Data of COVID-19 cases and vaccinations was acquired from CovidTracker and US Center for Disease Control and Prevention, respectively.
```{r}

```

Preliminary Results (provide summary statistics in tabular form and publication-quality figures, take a look at the kable function from knitr to write nice tables in Rmarkdown)

```{r}

```

Conclusion





